Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-02 analysis/paper/article.qmd
add plot showing self-reported user category counts, with color palette from "The Thing"

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 900
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
abstract: |
  Many archaeology graduate students pursue advanced degrees in the hope of undertaking an academic career. Job listing websites often serve as the first port-of-call for students  seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examine changes in the academic job market over time. We analyzed the text of 431 job ads posted from 2013--2023. Our analysis addresses the question of how  archaeological topics, methods, and geographic regions specified in archaeological job ads shifted over time. We also investigate whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? We assess the influence of socio-political factors on the changing focus of research topics in the field. This research aims to assist current and future archaeology students, graduates, and advisors to better understand the dynamics of the academic job market and the requirements of employers.
keywords: |
  archaeology; hiring; academia; job ads 
---

@@

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for  three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 Bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year."

@@

# visualisations ----------------------------------

# Number of edits per year
p_edits_per_year <-
ggplot(edits_for_each_year_tbl) +
  aes(as_date(edit_year)) +
  geom_bar() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# Number of unique editors per year
p_editors_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  summarise(n_distinct_editors = n_distinct(edit_name)) %>%
ggplot() +
  aes(as_date(edit_year),
      n_distinct_editors) +
  geom_col() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of editors\n(distinct usernames or IP addresses)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

# Most editors are only active for less that one year
p_editor_life_distr <-
ggplot(edits_for_each_year_tbl_span) +
  aes(edit_span_years) +
  geom_histogram() +
  theme_minimal() +
  scale_y_log10() +
  labs(x = "Years editing", y = "Number of editors")

@@

# for those editors active for more than three months:
p_editor_life_distr_year <-
edits_for_each_year_tbl_span %>%
 filter(edit_span_years > 0.25) %>%
ggplot() +
  aes(as_date(first_edit_year),
      reorder(edit_name, first_edit_year)) +
  geom_segment(aes(x =   first_edit,
                   xend = last_edit),
               linewidth = 1,
               lineend = "butt") +
  theme_minimal() +
  labs(x = "", y = "Individual editors\n(usernames and IP addresses redacted)") +
  theme(axis.text.y=element_blank())

# size of edits per year
p_edit_size_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  summarise(edit_size_num = sum(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Sum of edits (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# typical edit size per year
p_median_edit_size_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  # median, because sometimes there are 1-2 very large
  # page reformatting edits
  summarise(edit_size_num = median(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Median edit size (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

# edits per section for each year
library(ggbeeswarm)
p_edit_number_per_entry_per_year <-
ggplot(section_edit_tally_tbl) +
  aes(as_date(edit_year),
      group = as_date(edit_year),
      n) +
  geom_boxplot() +
  theme_minimal() +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits per job ad") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

library(cowplot)
plot_grid(
  plot_grid(p_edits_per_year,
          p_editors_per_year,
          p_median_edit_size_per_year,
          p_edit_number_per_entry_per_year,
          labels = c("A", "B", "D", "E")),
          plot_grid(p_editor_life_distr_year,
                    p_editor_life_distr,
                    rel_widths = c(2),
                    ncol = 1,
                    labels = c("C", "F"))
          )
```

@@

#  the resulting tidy table
tidy_table_to_plot <- 
tidy_table %>%
  filter(!status %in% c("I'm a lurker",
                        "I have a current but term-limited faculty job",
                        "I have some other teaching, research, or non-academic gig",
                        "Nope (independent scholar, on sabbatical from adjuncting, etc.)")) %>%
  mutate(year = str_replace(year, "-", "-\n"))

@@

  ggplot(tidy_table_to_plot) +
  aes(year,
      count) +
  geom_col() +
  theme_minimal() +
  facet_wrap(~ status,
             scales = "free") +
  xlab("") +
  theme(axis.text.x = element_text(size = 6))

tidy_table_to_plot %>% 
filter(status != "How many people use the wiki?") %>% 
ggplot() +
  aes(year,
      count,
      fill = status) +
  geom_col() +
  theme_minimal() +
  xlab("") +
  scale_fill_brewer(type = "qual") +
  theme(axis.text.x = element_text(size = 6),
        legend.position = c(.7,.7))
```

There seems to be a shift over time in the balance of editors in these two groups, as indicated by the number of self-reporting users over time. @fig-panel-wiki-self-report-numbers shows the data found on each page in a section labeled 'current users', where users can volunteer to update a table tracking the status of users. These data indicate that the number of self-reporting users, from the values recording at 'How many people use the wiki?', declined by `r how_many_people_use_the_wiki_perc_start_end`% during the study period. At the same time, the proportion of users self-identifying as "search committee members" increased, from under `r search_committee_members_perc_start`% in 2014-2018 to over `r search_committee_members_perc_end`% from 2020 onward. This suggests that over time fewer job applicants have been contributing job postings and sharing information about their experiences, and search committee members were increasingly posting job advertisements themselves. Overall the Archaeology Academic Jobs Wiki is best understood not as a comprehensive archive of all jobs posted during the study period, but a biased sample of the job ads that were of most interest to archaeologists applying for jobs in a given year. It represents the jobs that early career applicants were most focused on applying for, and getting information on.

>>>>

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
abstract: |
  Many archaeology graduate students pursue advanced degrees in the hope of undertaking an academic career. Job listing websites often serve as the first port-of-call for students  seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examine changes in the academic job market over time. We analyzed the text of 431 job ads posted from 2013--2023. Our analysis addresses the question of how  archaeological topics, methods, and geographic regions specified in archaeological job ads shifted over time. We also investigate whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? We assess the influence of socio-political factors on the changing focus of research topics in the field. This research aims to assist current and future archaeology students, graduates, and advisors to better understand the dynamics of the academic job market and the requirements of employers.
keywords: |
  archaeology; hiring; academia; job ads 
---

@@

@@

# visualisations ----------------------------------
small_base_size <- 4

# Number of edits per year
p_edits_per_year <-
ggplot(edits_for_each_year_tbl) +
  aes(as_date(edit_year)) +
  geom_bar() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# Number of unique editors per year
p_editors_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  summarise(n_distinct_editors = n_distinct(edit_name)) %>%
ggplot() +
  aes(as_date(edit_year),
      n_distinct_editors) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of editors\n(distinct usernames or IP addresses)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

# Most editors are only active for less that one year
p_editor_life_distr <-
ggplot(edits_for_each_year_tbl_span) +
  aes(edit_span_years) +
  geom_histogram() +
  theme_minimal(base_size = small_base_size) +
  scale_y_log10() +
  labs(x = "Years editing", y = "Number of editors")

@@

# for those editors active for more than three months:
p_editor_life_distr_year <-
edits_for_each_year_tbl_span %>%
 filter(edit_span_years > 0.25) %>%
ggplot() +
  aes(as_date(first_edit_year),
      reorder(edit_name, first_edit_year)) +
  geom_segment(aes(x =   first_edit,
                   xend = last_edit),
               linewidth = 1,
               lineend = "butt") +
  theme_minimal(base_size = small_base_size) +
  labs(x = "", y = "Individual editors\n(usernames and IP addresses redacted)") +
  theme(axis.text.y=element_blank())

# size of edits per year
p_edit_size_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  summarise(edit_size_num = sum(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Sum of edits (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# typical edit size per year
p_median_edit_size_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  # median, because sometimes there are 1-2 very large
  # page reformatting edits
  summarise(edit_size_num = median(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Median edit size (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

# edits per section for each year
library(ggbeeswarm)
p_edit_number_per_entry_per_year <-
ggplot(section_edit_tally_tbl) +
  aes(as_date(edit_year),
      group = as_date(edit_year),
      n) +
  geom_boxplot() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits per job ad") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))
```

@@

#  the resulting tidy table
tidy_table_to_plot <- 
tidy_table %>%
  filter(!status %in% c("I'm a lurker",
                        "I have a current but term-limited faculty job",
                        "I have some other teaching, research, or non-academic gig",
                        "Nope (independent scholar, on sabbatical from adjuncting, etc.)")) %>%
  mutate(year = str_sub(year, 6L))

@@

# "The Thing" Inspired Palette (8 Colors)
the_thing_palette <- c(
  "#6488AC", # 1. Icy Blue
  "#1A2A3A", # 2. Deep Shadow Blue/Black
  "#E8702A", # 3. Flamethrower Orange
  "#9D2323", # 4. Blood Red/Creature Tissue
  "#8B7355", # 5. Outpost Brown
  "#E1E1E1", # 6. Snowy Off-White/Grey
  "#5F7C6A", # 7. Parka Green/Drab
  "#D9A443"  # 8. Warning Yellow/Orange (Muted)
)

how_many_people_use_the_wiki <- 
tidy_table_to_plot %>% 
  filter(status != "How many people use the wiki?") %>% 
  mutate(status = str_wrap(status, 26)) %>% 
ggplot() +
  aes(year,
      count,
      fill = status) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  xlab("") +
  scale_fill_manual(values = the_thing_palette) +
  theme(legend.position = c(.7,.8),
        legend.key.size = unit(0.25,"line"),
        legend.text = element_text(margin = margin(r = 10, 
                                                   t = 2,
                                                   b = 2,
                                                   unit = "pt"))) +
  guides(fill=guide_legend(nrow=3,
                           byrow=TRUE),
         )
```

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 Bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year. G: Breakdown of self-reported Wiki page user categories for each year."
#| fig-height: 5

library(cowplot)
plot_grid(
 plot_grid(
  plot_grid(p_edits_per_year,
          p_editors_per_year,
          p_median_edit_size_per_year,
          p_edit_number_per_entry_per_year,
          labels = c("A", "B", "D", "E"),
          label_size = small_base_size),
          plot_grid(p_editor_life_distr_year,
                    p_editor_life_distr,
                    rel_widths = c(2),
                    ncol = 1,
                    labels = c("C", "F"),
                    label_size = small_base_size)),
  plot_grid(how_many_people_use_the_wiki,
            labels = "G",
            label_size = small_base_size),
  rel_heights = c(2,1),
  ncol = 1)
```

There seems to be a shift over time in the balance of editors in these two groups, as indicated by the number of self-reporting users over time. Panel G of @fig-panel-wiki-characteristics shows the data found on each page in a section labeled 'current users', where users can volunteer to update a table tracking the status of users. These data indicate that the total number of self-reporting users, from the values recording at 'How many people use the wiki?', declined by `r how_many_people_use_the_wiki_perc_start_end`% during the study period. At the same time, the proportion of users self-identifying as "search committee members" increased, from under `r search_committee_members_perc_start`% in 2014-2018 to over `r search_committee_members_perc_end`% from 2020 onward. This suggests that over time fewer job applicants have been contributing job postings and sharing information about their experiences, and search committee members were increasingly posting job advertisements themselves. Overall the Archaeology Academic Jobs Wiki is best understood not as a comprehensive archive of all jobs posted during the study period, but a biased sample of the job ads that were of most interest to archaeologists applying for jobs in a given year. It represents the jobs that early career applicants were most focused on applying for, and sharing information on.
